---
title: "multivariate_assignment"
author: "Julius Alipala"
date: "February 7, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
layout: page
title: multivariate models
---

For this assignment will be analyzing data on the Vegetation
and Environment in Dutch Dune Meadows.

To import the data and read the metadata run the following:

```{r}
library(vegan)
data(dune)
data(dune.env)
```

1. Conduct an indirect ordination on the dune plant community. Specifically,
visually examine a NMDS plot using the bray-curtis distance metric. Below is 
some code to help you develop a potential plot that emphasizes the role of the
environmental variable "Moisture". Describe how you interpret the
graphic. What is the goal of creating such a plot? Does this analysis suggest
any interesting findings with respect to the dune vegetation?


```{r}
dune_mds = metaMDS(dune)
```

```{r}
plot(dune_mds, type='n')
text(dune_mds, 'sp', cex=.5)
# generate vector of colors 
color_vect = rev(terrain.colors(6))[-1]
points(dune_mds, 'sites', pch=19, 
       col=color_vect[dune.env$Moisture])
legend('topright', paste("Moisture =", 1:5, sep=''), 
       col=color_vect, pch=19)

```

This plot shows that each site's position along the horizontal axis is dependent on it's moisture level. This indicates that moisture is a key factor in determining what species are found at a site. A species commonly found in a site with a certain moisture level has a good chance of appearing in another site with that same moisture level.

2. Carry out a direct ordination using CCA in order to test any potential 
hypotheses that you developed after examining the MDS plot. Specifically,
carry out a test of the entire model (i.e., including all constrained axes)
and also carry out tests at the scale of individual explanatory variables
you included in your model if you included more than one variable. Plot your 
results.

```{r}
dune_cca = cca(dune ~ ., data=dune.env)

RsquareAdj(dune_cca, 100)
anova(dune_cca, permutations=999)
anova(dune_cca, by='margin', permutations=999)
```

```{r}
plot(dune_cca, type='n', scaling=1)
text(dune_cca, display='sp', cex=0.5, scaling=1, col='blue')
points(dune_cca, 'sites', pch=19, 
           col=color_vect[dune.env$Moisture])
text(dune_cca, display='bp', col='red')
```
```{r}
dune_cca_m = cca(dune ~ Moisture, data=dune.env)

RsquareAdj(dune_cca_m, 100)
anova(dune_cca_m, permutations=999)
anova(dune_cca_m, by='margin', permutations=999)
```

```{r}
plot(dune_cca_m, type='n', scaling=1)
text(dune_cca_m, display='sp', cex=0.5, scaling=1, col='blue')
points(dune_cca_m, 'sites', pch=19, 
           col=color_vect[dune.env$Moisture])
text(dune_cca_m, display='bp', col='red')
```

dune ~ Moisture is the better model according to adjusted R^2.


3. Do your two analyses agree with one another or complement one another or do
these two analyses seem to be suggesting different take home messages? Which
analysis do you find to be more useful?

The plot of dune ~ Moisture shows that Moisture.L explains most of the variance along the horizontal axis. This agrees with the observations made during the NMDS analysis. I find the CCA analysis to be more useful since it can easily test hypotheses made during earlier analysis. 