---
title: "spatial_assigment"
author: "Julius Alipala"
date: "February 18, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spatial Modeling Assignment

```{r}
library(vegan)
library(nlme)
data(BCI)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
```

1) Examine if there is evidence of spatial dependence in a rare and a common
species in the BCI tree dataset

```{r}
# count of species
sp_count = colSums(BCI, na.rm=FALSE, dims=1)
summary(sp_count)

# boxplot of sp_count without outliers
boxplot(sp_count, outline = FALSE)

# rare species: <20
# return index of species with counts between 15-20
rare_sp = which(sp_count < 20 & sp_count > 15)
rare_sp

# chose Cavanillesia platanifoli to investigate further
# index of Cavanillesia platanifoli
cav_plat_index = 33
sprintf("count of Cavanillesia platanifoli: %i", sp_count[cav_plat_index])

# common species: > 150
# return index of species with counts between 150-200
comm_sp = which(sp_count > 150 & sp_count < 200)
comm_sp

# chose Alchornea costaricensis to investigate further
# index of Alchornea costaricensis
alc_cos_index = 7
sprintf("count of Alchornea costaricensis: %i", sp_count[alc_cos_index])

```

Based on the results above, I've chosen Cavanillesia platanifoli to represent my rare species and Alchornea costaricensis to represent my common species.

```{r}
# index of sites which contain Alchornea costaricensis
alc_cos_sites = which(BCI[alc_cos_index]>0)
# sites that have Alchornea costaricensis colored green; otherwise colored red
plot(BCI_xy, cex=2, pch=19, col=ifelse(rownames(BCI_xy) %in% alc_cos_sites,'green', 'red'))


```

The graph above shows sites which contain Alchornea costaricensis in green. Sites without Alchornea costaricensis are colored red. Based on this graph, there is no evidence of spatial dependence for the common species, Alchornea costaricensis. 

```{r}
# index of sites which contain Cavanillesia platanifoli
cav_plat_sites = which(BCI[cav_plat_index]>0)
# sites that have Cavanillesia platanifoli colored green; otherwise colored red
plot(BCI_xy, cex=2, pch=19, col=ifelse(rownames(BCI_xy) %in% cav_plat_sites,'green', 'red'))


```

The graph above shows evidence of spatial dependence for the rare species, Cavanillesia platanifoli. The sites that contain this species are grouped in a certain area and are all adjacent to each other. No solitary sites contain the species.

2) Build two generalized linear models to predict the abundance of the species
*Drypetes standleyi* using the abundance of other tree species in the study site.
Specifically examine the following species as predictor variables:

```{r}
sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")
x = BCI_xy$x
y = BCI_xy$y
```
Note renaming the species ids to something a little easier to work with like
"sp_a", "sp_b" will make model construction a little less cumbersome

* Model 1: only include a single species as a predictor variable
```{r}
ds_lm_single = gls(Drypetes.standleyi ~ Cordia.lasiocalyx, data=BCI)
par(mfrow=c(1,1))
plot(Variogram(ds_lm_single, form = ~ x + y))
  

```


* Model 2: include all of the species as predictor variables
```{r}
ds_lm_all = gls(Drypetes.standleyi ~ Cordia.lasiocalyx + Hirtella.triandra + Picramnia.latifolia + 
              Quassia.amara + Tabernaemontana.arborea + Trattinnickia.aspera +
              Xylopia.macrantha, data=BCI)
par(mfrow=c(1,1))
plot(Variogram(ds_lm_all, form = ~ x + y))
  

```

With both models examine the spatial dependence of the residuals using the 
function `Variogram`. Model the spatial dependence in the residuals using one
of the error structures available. 

```{r}
xy_dist = dist(BCI_xy)
max_dist = max(xy_dist)/2
# rational quadratic
ds_single_rat = update(ds_lm_single, corr=corRatio(form=~x+y))
plot(Variogram(ds_single_rat, maxDist = max_dist))
plot(Variogram(ds_single_rat, resType='normalized', maxDist = max_dist))
```

The rational quadratic model does not seem to fit very well, but the plot of residuals shows a better fit.

```{r}
# gaussian
ds_all_gauss = update(ds_lm_all, cor=corGaus(form=~x+y))
plot(Variogram(ds_all_gauss, maxDist = max_dist))
plot(Variogram(ds_all_gauss, resType='normalized', maxDist=max_dist))
```

The Gaussian model does not seem to fit well at all. I would expect the plot for residuals to drop a little more as the distance increases.


* Did including the spatial error term have a large impact on the coefficients
of the model?
```{r}
summary(ds_lm_single)$coefficients
summary(ds_single_rat)$coefficients

```
Spatial error term has a large impact on single species predictor.


```{r}
summary(ds_lm_all)$coefficients
summary(ds_all_gauss)$coefficients
```

In general, spatial error term did not change the coefficients much for the models using 7 species as a predictor.

* Did including the spatial error terms significantly improve model fit (use  function `anova` to carry out model comparison)?
```{r}
# compare single species models
anova(ds_lm_single, ds_single_rat)
```
Spatial error terms improved the model for the single species as a predictor. I expected the model without spatial error terms to be a better fit from looking at the plots.


```{r}
# compare all species models
anova(ds_lm_all, ds_all_gauss)
```
No significant improvements were made by including spatial error terms for the model with 7 species as a predictor.

* Explain why you did or did not observe a difference in the influence of adding the spatial error term between the two models.

For the model with a single species as a predictor, including spatial error terms showed a significant improvement in fit. Since only one species is used as a predictor, that species' features are the only variables driving the prediction. If the predictor species is very different from the target species, this may result in a large difference between predicted and actual data.

Using multiple species as a predictor may provide a more generalized prediction. This is why including spatial error terms has little impact on the model using multiple species as predictors.
